Validation of Hepatitis B viral count at ICL
========================================================

These are the results of the validation of the Hepatitis B viral count at Internatinal Clinical Laboratories, Addis Abeba. 
Serum specimens were run on the Abbott m2000 system and results were compared with results of the same specimens obtained from Bioscientia (Germany).

```{r,echo=FALSE, results='hide'}
setwd("~/Documents/workspace/R-project/HBV")

# libraries
library(ggplot2)
library(corrgram)
library(MethComp)

# read file
HepB_ICL <- read.delim("~/Documents/workspace/R-project/HBV/HepB_ICL.txt")
```
In a summary. 'Zero' values have been changed to '1' in order to be able to plot in logaritmic scale. The lower limit of detection (LLD) at ICL is 10 IU/ml and the LLD at Bioscientie os 20IU/ml. So, if the result is <20IU/ml, the detected value could be anywhere between 1 and 20. Therefore, the lower limit of detection has been set for ICL at '5 IU/ml' and '10 IU/ml' for Bioscientia. 

```{r Summary}
summary(HepB_ICL)
head(HepB_ICL)
```
To make it more easy, the set of values from Bioscientia = 'x'. The set of values from ICL = 'y' 
```{r,echo=FALSE}
x <- HepB_ICL$Bioscientia
y <- HepB_ICL$ICL
```
Calculate the means and difference between the two sets (x and y)
```{r Mean}
# derive difference
mean(x)
mean(y)
# mean Bio - mean ICL
mean(x)-mean(y)
```
Because n=17 is small, the distribution of the differences should be approximately normal. Check using a boxplot and QQ plot. There is some skew.
```{r Boxplot}
HepB_ICL$diff <- x-y
# HepB_ICL$diff

boxplot(HepB_ICL$diff)
qqnorm(HepB_ICL$diff)
qqline(HepB_ICL$diff)
```
Shaphiro test of normality. 
```{r Shapiro}
shapiro.test(HepB_ICL$diff)
```
The normality test gives p < 0.003, which is small, so we reject the null hypothesis that the values are distributed normally.

This means that we cannot use the student t-test. Instead, use the Mann-Whitney-Wilcoxon Test, we can decide whether the population distributions are identical without assuming them to follow the normal distribution.
```{r Wilcoxon}
wilcox.test(x, y, paired = TRUE)
```
p > 0.05 and therefore the H0 is NOT rejected. The two populations are identical.

Just to see what happens in the Student T-test.
A paired t-test: one sample, two tests
H0 = no difference; H1 = mean of 2 tests are different
mu= a number indicating the true value of the mean 
(or difference in means if you are performing a two sample test).
```{r t-test}
t.test(x, y, mu=0, paired=T, alternative="greater")
```
p = 0.756. Because p is larger than alpha, we do NOT reject H0.
In other words, it is unlikely the observed agreements happened by chance. 
However, because the populations do not have a normal distribution, we can not use the outcome if this test.

For correlation, three methods are used: pearson, kendall and spearman at a confidence level of 95%.
```{r Correlation}
# correlation of the two methods
cor.test(x, y, 
         alternative = c("two.sided", "less", "greater"),
         method = c("pearson", "kendall", "spearman"),
         exact = NULL, conf.level = 0.95)
```
The correlation with the spearman test is 0.9398081. Almost perfect correlation. 
```{r,echo=FALSE,results="hide"}
# drop first column
HepB_ICL$PIN <- NULL
```
Plotting the two methods using logarithmic scales.
```{r Plotting}
g <- ggplot(HepB_ICL, aes(log(ICL), log(Bioscientia)))


# add layers
g + 
  geom_smooth(method="lm", se=TRUE, col="steelblue", size = 1) +
  geom_point(size = 3, aes(colour = x)) +
  scale_colour_gradient("IU/ml", high = "red", low = "blue", space = "Lab") +
  labs(y = "BioScientia (log IU/ml)") +
  labs(x = "ICL (log IU/ml)") +
  theme_bw(base_family = "Helvetica", base_size = 14) +
  scale_x_continuous(breaks=c(0,4,8,12))
```
Summary data on the correlation line.
```{r}
regmod <- lm(y~x, data=HepB_ICL)
summary(regmod)
```

The Bland-Altman Analysis. To check if there is a bias.
```{r, echo=FALSE}
# convert table to datafram
as.data.frame.matrix(HepB_ICL) 
```
```{r}
BlandAltman(x, y,
            x.name = "BioScientia IU/ml",
            y.name = "ICL IU/ml",
            maintit = "Bland-Altman plot for HBV count",
            cex = 1,
            pch = 16,
            col.points = "black",
            col.lines = "blue",
            limx = NULL,
            limy = NULL,
            ymax = NULL,
            eqax = FALSE,
            xlab = NULL,
            ylab = NULL,
            print = TRUE,
            reg.line = FALSE,
            digits = 2,
            mult = FALSE)
```
When the dots are around 0, the two test could be interchanged for a patient. The two test can be interchanged. 



