---
output:
  pdf_document: default
  html_document: default
---
Validation of Hepatitis B viral count using R-computing
========================================================

This is an example on how R-computing can be used for validation of an quantitative assay. In this case two assays for Hepatitis B viral count are compared. 

```{r,echo=FALSE, results='hide'}
setwd("~/Documents/workspace/R-project/HBV")

# libraries
library(ggplot2)
library(corrgram)
library(MethComp)

# read file
HepB_Web <- read.delim("~/Documents/workspace/R-project/HBV/HepB_Web.txt")
head(HepB_Web)
```
In a summary. 'Zero' values have been changed to '1' in order to be able to plot in logaritmic scale. The lower limit of detection (LLD) at home-lab is 10 IU/ml and the LLD at the reference-lab os 20 IU/ml. So, if the result is <20IU/ml, the detected value could be anywhere between 1 and 20. Therefore, the lower limit of detection has been set for home-lab at '5 IU/ml' and '10 IU/ml' for the reference lab. 

```{r Summary}
summary(HepB_Web)
head(HepB_Web)
```
To make it more easy, the set of values from Reference-lab = 'x'. The set of values from Home-lab = 'y' 
```{r,echo=FALSE}
x <- HepB_Web$Ref_lab
y <- HepB_Web$Home_lab
```
Calculate the means and difference between the two sets (x and y)
```{r Mean}
# derive difference
mean(x)
mean(y)
# mean Ref_lab - mean Home_lab
mean(x)-mean(y)
```
Because n=17 is small, the distribution of the differences should be approximately normal. Check using a boxplot and QQ plot. There is some skew.
```{r Boxplot}
HepB_Web$diff <- x-y
HepB_Web$diff

boxplot(HepB_Web$diff)
qqnorm(HepB_Web$diff)
qqline(HepB_Web$diff)
```
Shaphiro test of normality. 
```{r Shapiro}
shapiro.test(HepB_Web$diff)
```
The normality test gives p < 0.003, which is small, so we 
reject the null hypothesis that the values are distributed normally. 

This means that we cannot use the student t-test. Instead, use the Mann-Whitney-Wilcoxon Test, we can decide whether the population distributions are identical without assuming them to follow the normal distribution.
```{r Wilcoxon}
wilcox.test(x, y, paired = TRUE)
```
p > 0.05 and therefore the H0 is NOT rejected. 
The two populations are identical.

Just to see what happens in the Student T-test.
A paired t-test: one sample, two tests
H0 = no difference; H1 = mean of 2 tests are different
mu= a number indicating the true value of the mean 
(or difference in means if you are performing a two sample test).
```{r t-test}
t.test(x, y, mu=0, paired=T, alternative="greater")
```
p = 0.759. Because p is larger than alpha, we do NOT reject H0.
In other words, it is unlikely the observed agreements happened by chance. 
However, because the populations do not have a normal distribution, we can not use the outcome if this test.

For correlation, three methods are used: pearson, kendall and spearman at a confidence level of 95%.
```{r Correlation}
# correlation of the two methods
cor.test(x, y, 
         alternative = c("two.sided", "less", "greater"),
         method = c("pearson", "kendall", "spearman"),
         exact = NULL, conf.level = 0.95)
```
The correlation with the spearman test is 0.9416175. Almost perfect correlation. 
```{r,echo=FALSE,results="hide"}
# drop first column
HepB_Web$PIN <- NULL
```
Plotting the two methods using logarithmic scales.
```{r Plotting}
g <- ggplot(HepB_Web, aes(log(Home_lab), log(Ref_lab)))

# add layers
g + 
  geom_smooth(method="lm", se=TRUE, col="steelblue", size = 1) +
  geom_point(size = 3, aes(colour = x)) +
  scale_colour_gradient("IU/ml", high = "red", low = "blue", space = "Lab") +
  labs(y = "Reference lab (log IU/ml)") +
  labs(x = "Home lab (log IU/ml)") +
  theme_bw(base_family = "Helvetica", base_size = 14) +
  scale_x_continuous(breaks=c(0,4,8,12))
```
Summary data on the correlation line.
```{r}
regmod <- lm(y~x, data=HepB_Web)
summary(regmod)
```

The Bland-Altman Analysis. To check if there is a bias.
```{r, echo=FALSE}
# convert table to datafram
as.data.frame.matrix(HepB_Web)
```
```{r}
BlandAltman(x, y,
            x.name = "Reference lab IU/ml",
            y.name = "Home lab IU/ml",
            maintit = "Bland-Altman plot for HBV count",
            cex = 1,
            pch = 16,
            col.points = "black",
            col.lines = "blue",
            limx = NULL,
            limy = NULL,
            ymax = NULL,
            eqax = FALSE,
            xlab = NULL,
            ylab = NULL,
            print = TRUE,
            reg.line = FALSE,
            digits = 2,
            mult = FALSE)
```
When the dots are around 0, the two test could be interchanged for a patient. So, the two test can be interchanged. There are, however, some outliners: large difference of viral count between the two labs.  



